---
# PLEASE SEE THE README for in depth description github.com/brentthorne/posterdown
# build with pagedown::chrome_print("metabolomics-2020-poster.Rmd")
poster_height: "48in"
poster_width: "36in"
font_family: 'Open Sans'
#ESSENTIALS
title: '**_Spectra_: A scalable and flexible infrastructure for mass spectrometry data in R**'
author:
  - name: '**Johannes Rainer**'
    affil: 1,*
    main: true
    orcid: '0000-0002-6977-7147'
    twitter: jo_rainer
    email: johannes.rainer@eurac.edu
  - name: Sebastian Gibb
    affil: 2
    orcid: '0000-0001-7406-4443'
  - name: Laurent Gatto
    affil: 3
    orcid: '0000-0002-1520-2268'
affiliation:
  - num: 1
    address: Institute for Biomedicine, Eurac Research, Affiliated Institute of the University of Lübeck, 39100 Bolzano, Italy.
  - num: 2
    address: Department of Anaesthesiology and Intensive Care, University Medicine Greifswald, University of Greifswald, 17475 Greifswald, Germany.
  - num: 3
    address: Computational Biology Unit, de Duve Institute, Université catholique de Louvain, Brussels, 1200, Belgium.
# Title Box
# colors:
# #cccdcc: light grey
# #0b4545: Eurac red
# #838b93: Eurac dark grey
# #1e2b38: darkest grey
titlebox_bordercol: "#cccdcc"
title_textcol: "#d04c29f"
author_textcol: "#212121"
title_textsize: "80pt"
author_textsize: "50pt"
authorextra_textsize: "30pt"
affiliation_textsize: "25pt"
affiliation_textcol: '#00000080'
caption_fontsize: "20pt"
logoleft_name: "images/Spectra.png"
logocenter_name: ""
logoright_name: "images/Eurac_Research.png"
# POSTER BODY OPTIONS
primary_colour: '#212121'
secondary_colour: '#d04c29f'
accent_colour: "#313131"
body_bgcol: "#ffffff"
body_textsize: "45px"
body_textcol: "#1e2b38"
column_numbers: 2
column_margins: "2cm"
columnline_col: "#cccdcc"
columnline_width: "1mm"
columnline_style: "dashed"
sectitle_textcol: "#d04c29"
sectitle_textsize: "50pt"
sectitle2_textsize: "40pt"
sectitle_bgcol: "#ffffff"
sectitle_bordercol: "#000000"
sectitle_borderwidth: "0mm"
reference_textsize: "23px"
code_fontfamily: 'Inconsolata'
#--Standard Options--#
output: 
  posterdown::posterdown_html:
    self_contained: false
    pandoc_args: --mathjax
    number_sections: false
    template: template_jo.html
bibliography: references.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = FALSE,
  warning = FALSE,
  message = FALSE,
  tidy = FALSE,
  fig.align = 'center'
)
```

# Introduction

- Data analysis/manipulation and storage functionality is separated into:
  - `Spectra`: provides functions to handle and analyze MS data.
  - `MsBackend`: manages and provides the MS data to `Spectra`.
- This simplifies extension of the package's functionality.
- **Use case**: match experimental MS2 spectra against public database (full version of the tutorial:
  [https://jorainer.github.io/SpectraTutorials/](https://jorainer.github.io/SpectraTutorials/)).

```{r libs, echo = FALSE}
library(Spectra)
library(MsBackendMgf)
library(MsBackendHmdb)
library(pheatmap)
library(MsCoreUtils)
```

# Import from mzML files

- Import data from an LC-MS/MS experiment (4 mzML files).
- `MsBackendMzR` supports data import from mzML/mzXML/CDF files; has a small
  memory footprint hence enabling analysis of large scale experiments.

```{r mzml-import}
fls <- dir("data/mzML", pattern = "mzML$", full.names = TRUE)
sps_all <- Spectra(fls, backend = MsBackendMzR())
```

- Identify MS2 spectra with precursor m/z matching the [M+H]+ ion of Cystine.

```{r}
mz <- 241.0311
sps <- filterPrecursorMz(sps_all, mz = mz + ppm(c(-mz, mz), 10))
```

- Plot first spectrum: raw spectra seem to be noisy.

```{r eval = FALSE}
plotSpectra(sps[1])
```

```{r raw-ms2, echo = FALSE}
par(mar = c(4, 4.2, 1.4, 0.5))
plotSpectra(sps[1])
```

- Use `filterIntensity` to remove intensities below 5% of base peak signal.
- *Normalize* each spectrum by applying a custom function `norm_int` to each
  spectrum with `addProcessing`.

```{r filterintensity}
low_int <- function(x) x > max(x) * 0.05
sps <- filterIntensity(sps, intensity = low_int)

norm_int <- function(x, ...) {
    x[, "intensity"] <- 100 * x[, "intensity"] / max(x[, "intensity"])
    x
}
sps <- addProcessing(sps, norm_int)
```

- Calculate pairwise similarity between spectra and visualize.

```{r sim}
cormat <- compareSpectra(sps, ppm = 20, FUN = ndotproduct)
hm <- pheatmap(cormat, cutree_rows = 3)
```

- Spectra group into 3 clusters: related to the applied collision energy.

- Proceed analysis with spectra from 20eV collision energy.

```{r ce20}
sps_ce20 <- split(sps, cutree(hm$tree_row, 3))[[1L]]
```

# Comparison against spectra from HMDB

- Next step: compare spectra against *reference* spectra from
  [HMDB](https://www.hmdb.ca).
- `MsBackendHmdbXml` supports import from HMDB MS/MS spectra xml files.

```{r hmdb-import, eval = FALSE}
library(MsBackendHmdb)
fls <- dir("data/hmdb_all_spectra/", full.names = TRUE, pattern = "ms_ms")
hmdb <- Spectra(fls, source = MsBackend)
```

```{r echo = FALSE}
load("data/hmdb.RData")
```

- Subset the ~ 500,000 spectra to those containing the precursor m/z.

```{r has-mz}
has_mz <- containsMz(hmdb, mz = mz, tolerance = 0.2)
hmdb_with_mz <- hmdb[has_mz]
```

- Compare the experimental spectra against the database.

```{r}
res <- compareSpectra(sps_ce20, hmdb_with_mz, tolerance = 0.2)
```

- Highest similarity is `r format(max(res), digits = 3)`. Plotting best matching
  spectra.

```{r, eval = FALSE}
idx <- which(res == max(res), arr.ind = TRUE)
label_fun <- function(x) format(unlist(mz(x)), digits = 4)
plotSpectraMirror(sps_ce20[idx[1]], hmdb_with_mz[idx[2]], tolerance = 0.2,
                  labels = label_fun, labelPos = 2, labelSrt = -30)
```

```{r, mirror-plot, echo = FALSE, fig.width = 10, fig.height = 6}
par(mar = c(3.5, 4.2, 1.4, 0.5))
idx <- which(res == max(res), arr.ind = TRUE)
label_fun <- function(x) format(unlist(mz(x)), digits = 4)
plotSpectraMirror(sps_ce20[idx[1]], hmdb_with_mz[idx[2]], tolerance = 0.2,
                  labels = label_fun, labelPos = 2, labelSrt = -30)
```


- Best match is with [HMDB0000192](https://hmdb.ca/metabolites/HMDB0000192)
  (L-Cystine).


# Export in MGF format

- Add annotation and collision energy to the spectra.
- Export to a file in [mascot generic
format](https://www.matrixscience.com/help/data_file_help.html) (MGF) using
`MsBackendMgf`.

```{r}
sps_ce20$hmdb_id <- hmdb_with_mz[idx[2]]$compound_id
sps_ce20$collisionEnergy <- 20

library(MsBackendMgf)
export(sps_ce20, backend = MsBackendMgf(), file = "Cystin_ce20.mgf")
```


# Conclusion and Outlook

- `Spectra` provides a flexible and expandable infrastructure for MS data in R.
- Enables seamless integration of MS data from different data sources or
  formats.
- Allows elegant MS data handling and analysis in R.
- Future backends will involve storage of data in SQL databases with possibility
  of remote access and eventually access to online spectral databases.


_Get or cite this poster here: [https://doi.org/10.5281/zenodo.4080896](https://doi.org/10.5281/zenodo.4080896)_
